### 2022-08-03 doc/VerticalArrangements.tex
This version incorporates Gloria's 6/3 edits to the document

\begin{abstract}
\noindent We use a vertical supply-chain model to study the welfare impacts of mergers with both horizontal and vertical aspects.  The model features logit Bertrand competition downstream and Nash Bargaining upstream.  We numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream horizontal/vertical mergers between an unintegrated retailer and an integrated retailer/wholesaler, (3) upstream horizontal/vertical mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs. We find that in for mergers with both horizontal and vertical aspects, consumers are typically net harmed.  
\end{abstract}

@@

The horizontal versus vertical distinction has been enshrined in U.S. Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The FTC withdrew its approval from the Vertical Merger Guidelines in 2021.  The DOJ did not.  }  Given these policy documents, it is not surprising that the agencies typically try to frame most mergers as either horizontal or vertical.  However, as part of the 2022 effort by the U.S. Department of Justice (DOJ) and the Federal Trade Commission (FTC) to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically''\footnote{The full question is 1.g. in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.  The text of the question reads: ``Should the guidelines' traditional distinctions between horizontal and vertical mergers be revisited in light of recent economic trends in the modern economy? What aspects of modern market realities may be lost by focusing on these relationships categorically? Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition? If so, how?''.  }  In this paper, our goal is to provide guidance to practitioners on how these mergers can be simulated in a familiar framework and to also show the range of welfare outcomes that are likely given this modeling environment.

We use a standard merger simulation model to assess the welfare implications of these complex mergers.  Our model is drawn directly from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.  We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.  In each of these simulations, we calculate the impacts on consumer welfare and firm profits.  Furthermore, we examine how the presence of additional integrated rivals (separate from the merging firms) changes our results.

Our simulations show certain patterns across mergers.  For mixed upstream and mixed downstream mergers, we find that consumers are net harmed in most instances.  The exceptions occur when upstream firms have more bargaining power than downstream firms, which is when we expect pre-merger input prices to be high and the gains from EDM to be the largest.  Furthermore, we find that integrated mergers are typically net harmful to consumers as well.  The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.  Thus, although all of these mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.  Furthermore, we find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.  The existence of rivals with their own sources of bargaining leverage stemming from vertical integration appears to reduce the variance of merger outcomes.

@@

The paper proceeds as follows.  In Section \ref{sec:theory}, we introduce the theoretical framework and explain how we model mergers in this framework.  Section \ref{sec:sims} explains how we construct our numerical simulations and provides their results, and our application to Republic/Santek appears in Section \ref{sec:application}.  Additional simulation results and the details of the data we use on the solid waste management markets appear in the Appendix.  Section \ref{sec:concl} concludes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Downstream and Upstream Competition}
Assume there is a set of consumers who each choose to buy a product sold by a single retailer.  Retailers are indexed by $r$, while the wholesalers that supply these retailers are indexed by $w$.  Prior to any mergers taking place, each wholesaler offers only one product, although each retailer can purchase from multiple wholesalers.  We denote the set of all retailers by $\mathbb{R}=\{1, \dots, \left\vert{\mathbb{R}}\right\vert\}$, and the set of all wholesalers by $\mathbb{W}=\{1, \dots, \left\vert{\mathbb{W}}\right\vert\}$.  We divide the set $\mathbb{W}$ into overlapping subsets, each labeled $\mathbb{W}^r$, to indicate which wholesalers' products are carried by which retailers.  Similarly, we divide the set of retailers $\mathbb{R}$ into overlapping subsets, each labeled $\mathbb{R}^w$, to indicate the retailers that carry the product sold by each wholesaler.

@@

Together the series of downstream and upstream first order conditions determine market equilibrium.  This model can be solved and calibrated as described in \citeN{ST2021}.  In our baseline configuration, we assume that wholesale and retail marginal costs are constant.  We loosen this restriction to allow for linear increasing marginal cost as a robustness check that appears in the Appendix.

@@

When the merged firm is bargaining with the unaffiliated wholesaler $v$ over what input price to pay as a retailer, the bargaining first order condition becomes
\begin{equation}
\begin{split}
&[p^W_{rv} - c^W_{rv}]s_{rv} - \sum_{t \in \mathbb{R}^v \setminus \{r\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^r \setminus \{v\}) = \\
&\frac{1-\lambda}{\lambda} \left([p_{rv} - p^W_{rv} - c^R_{rv}]s_{rv} - \sum_{x \in \mathbb{W}^r \setminus \{w,v\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{v\})\right.\\
&\left.- \underbrace{[p_{rw} - c^W_{rw} - c^R_{rw}] \Delta s_{rw}(\mathbb{W}^r \setminus \{v\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{wholesale recapture leverage effect}}\right).
\end{split}
\label{eq: vmerger ret upstream foc}
\end{equation}
In this case, the merged firm has two channels for potential additional profits should it cease to trade with wholesaler $v$.  First, if retail sales are diverted to product $w$ sold by retailer $r$, those sales will earn a higher margin due to lower marginal costs stemming from what we call the ``EDM recapture effect.''  Second, the loss of product $v$ carried by retailer $r$ could increase sales by wholesaler $w$ through other retailers, which we call the ``wholesale recapture leverage effect.''  Both of these effects would tend to lower the resulting input price.

@@

Turning to wholesale prices, again consider a merger between a preexisting integrated retailer/wholesaler $rw$ and a standalone retailer $s$.  When wholesaler $w$ bargains with an unaffiliated retailer $u$ we have the first order condition given by,
\begin{equation}
\begin{split}
&[p^W_{uw} - c^W_{uw}]s_{uw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s,u\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^u \setminus \{w\})\\
& - \overbrace{\sum_{t \in \{r,s\}}\left(\overbrace{[p_{tw}-c^W_{tw}-c^R_{tw}] \Delta s_{tw}(\mathbb{W}^u \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^t \setminus \{w\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathbb{W}^u \setminus \{w\})\right)}^{\text{RRC effect}}=\\
&\frac{1-\lambda}{\lambda} \left([p_{uw} - p^W_{uw} - c^R_{uw}]s_{uw} - \sum_{x \in \mathbb{W}^u \setminus \{w\}} [p_{ux} - p^W_{ux} - c^R_{ux}] \Delta s_{ux}(\mathbb{W}^u \setminus \{w\})\right).
\end{split}
\label{eq: intdmerger wh upstream foc}
\end{equation}
The RRC effect is augmented with the profits emanating from the sales of retailer $s$, in addition to the sales of retailer $r$, both of which may potentially recapture sales should retailer $u$ lose access to product $w$.  The merged firm has higher bargaining leverage as a result.  If instead the integrated retailer/wholesaler $rw$ were to merge with a standalone wholesaler $v$, then the profits earned by firm $rw$ should the negotiation with retailer $u$ fail are augmented with the earnings of wholesaler $v$ rather than of retailer $s$.  This adds a term similar to the wholesale recapture leverage effect seen in equation \eqref{eq: vmerger ret upstream foc} to the left-hand side of the bargaining first order condition.\footnote{For simplicity, we assume that when a retailer fails to reach an agreement with wholesaler $w$, that retailer's contract with wholesaler $v$ remains in place.  This assumption can be loosened.  }    In the case of a merger between two integrated retailer/wholesaler firms, all of these effects would appear.

Returning to a merger between an integrated retailer/wholesaler $rw$ and a standalone retailer $s$, when the merged firm bargains with an unaffiliated wholesaler $v$ to supply retailer $r$, the first order condition becomes
\begin{equation}
\begin{split}
&[p^W_{rv} - c^W_{rv}]s_{rv} - \sum_{t \in \mathbb{R}^v \setminus \{r\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^r \setminus \{v\}) = \\
&\frac{1-\lambda}{\lambda} \left([p_{rv} - p^W_{rv} - c^R_{rv}]s_{rv} - \sum_{x \in \mathbb{W}^r \setminus \{w,v\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{v\})\right.\\
&- \underbrace{\sum_{t \in \{r,s\}}[p_{tw} - c^W_{tw} - c^R_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{EDM recapture effect}}
- \underbrace{\sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^r \setminus \{v\})}_{\text{retail recapture leverage effect}}\\
&-\underbrace{\left.\sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})\right)}_{\text{wholesale recapture leverage effect}}.
\end{split}
\label{eq: intdmerger ret upstream foc}
\end{equation}
Compared to equation \eqref{eq: vmerger ret upstream foc}, now the EDM recapture effect applies to sales of product $w$ at both affiliated retailers $r$ and $s$.  Furthermore, there is also the possibility that sales will be diverted to retailer $s$ should retailer $r$ lose access to product $v$, which creates an additional ``retail recapture leverage effect'' alongside the wholesale recapture leverage effect.  All of these additional terms tend to increase the bargaining leverage of the merged firm.  If we instead examined a merger between retailer/wholesaler $rw$ and an unaffiliated wholesaler, that would augment the EDM recapture effect with sales of retailer $r$ for two merged wholesalers, rather than sales through two merged retailers.  Similarly, the retail recapture leverage effect would be replaced with an additional wholesale recapture leverage effect for the newly merged wholesaler.  For a merger between two integrated retailer/wholesaler firms, all of these effects would enter.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Simulations \label{sec:sims}}
In this section, we use a similar simulation setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations and also allow for the presence of integrated rival firms.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Generating Process}
Broadly speaking, we consider four categories of mergers: downstream, upstream, vertical, and integrated.  We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.  Similarly, upstream mergers are those between two unintegrated wholesalers or between an unintegrated wholesaler and a vertically integrated wholesaler/retailer.  We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.  Finally, we assume that an integrated merger is between two firms that are both already vertically integrated pre-merger.

For downstream, upstream, and vertical mergers, we simulate markets by randomly sampling shares from a Dirichlet distribution for 2 to 5 retailers or wholesalers, respectively.\footnote{We parameterize the Dirichlet distribution so it is equivalent to a uniform distribution.  }  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers in our simulated markets to 7. We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers). Vertically integrated firms are not siloed: integrated wholesalers supply inputs to retailers other than their integrated partner and vertically integrated retailers purchase inputs from wholesalers other than their integrated partner.

@@

The bargaining parameter ranges from from 0.1 (where wholesalers have the advantage) to 0.9 (where retailers have the advantage). We report our results in terms of relative bargaining power, $(1-\lambda)/\lambda$, which ranges from $9$ (wholesaler power is nine times greater than retailer power) to $1/9$ (retailer power is nine times greater than wholesaler power).  The bargaining parameter is identical for all of the retailers in each simulation.  We calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option.  All other goods are differenced relative to this option, resulting in the mean zero outside good normalization.  We set the market size to 1.

@@

Table \ref{tab:simsum} provides summary statistics for our simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.}  For reference, the 2010 DOJ/FTC Horizontal Merger Guidelines categorize ``Highly Concentrated Markets'' as those with HHIs over 2,500 points.  The Guidelines state that mergers with HHI changes greater than 200 points that result in Highly Concentrated markets are ``presumed likely to enhance market power.''\footnote{See \S 5.3 of the Guidelines.  }

The median average wholesale pre-merger price is \$5.80, and the median average retail pre-merger price is \$15.  Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.  Pre-merger HHIs range from 2,911 at the $25^{th}$ percentile to 4,981 at the $75^{th}$, with a median of 3,747. HHIs for vertical mergers increase by 1,328 points at the median, resulting in a median post-merger HHI equal to 6,048.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.  }  HHIs for downstream mergers increase by 2,128 points at the median, resulting in a median post-merger HHI equal to 5,153.  HHIs for upstream mergers increase by 2,070 points at the median, resulting in a median post-merger HHI equal to 4,968. HHIs for integrated mergers increase by 2,803 points at the median, resulting in a median post-merger HHI equal to 6,998.

Figure \ref{fig:surplussum} summarizes our results for welfare.  Each of the four panels show how mergers impact surplus for a particular set of agents (consumers, retailers, wholesalers, or the entire market combined, respectively).  Surplus is reported as a percentage change of total pre-merger expenditures in the downstream market.  Each panel contains four box and whisker plots, with each plot corresponding to a different type of merger. The whiskers show the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median. Negative outcome values imply harm, and positive values imply benefits.

@@

Throughout these simulations, we maintain the assumption that marginal costs are constant.  However, given that many of the mergers we study have vertical aspects where input cost considerations may drive the realization of EDM and RRC, it is reasonable to wonder how our results would change in an alternative marginal cost environment.  We explore the robustness of our results to the constant cost assumption in the Appendix, where we provide results that allow marginal costs to increase linearly.  Our simulations show that having linearly increasing marginal costs tends to result in mergers that are more harmful.

@@

Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.  Figure \ref{fig:CVvertincumbBW_consumer} depicts box and whisker plots summarizing consumer welfare as the number of incumbent integrated firms increases. For the results on total welfare, see the Appendix.  The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated. By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties. For integrated mergers, the plot when the number of incumbent integrated firms equals 2 (the lowest number allowed in this instance) depicts a merger between two integrated firms, again with no integrated rivals.

@@

\begin{figure}
\centering
\includegraphics{output/TrashSimsCompare.png}
\caption{Merger Simulation Results Comparison}
\label{fig:TrashSimsCompare}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Robustness to Increasing Marginal Costs}

>>>>

\begin{abstract}
\noindent We study the welfare impacts of mergers with both horizontal and vertical aspects.  Our model features logit Bertrand competition downstream and Nash Bargaining upstream.  We numerically simulate four merger types: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and (4) integrated mergers between two integrated retailer/wholesaler pairs. We find that mergers with both horizontal and vertical characteristics typically harm consumers.  We also apply the model to the Republic/Santek merger as a real-world example.  
\end{abstract}

@@

The horizontal versus vertical distinction has been enshrined in U.S. Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The FTC withdrew its approval from the Vertical Merger Guidelines in 2021.  The DOJ did not.  }  Given these policy documents, it is not surprising that the agencies typically try to frame most mergers as either horizontal or vertical.  However, as part of the 2022 effort by the DOJ and the FTC to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically''\footnote{The full question is 1.g. in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.  The text of the question reads: ``Should the guidelines' traditional distinctions between horizontal and vertical mergers be revisited in light of recent economic trends in the modern economy? What aspects of modern market realities may be lost by focusing on these relationships categorically? Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition? If so, how?''.  }  In this paper, our goal is to provide guidance to practitioners on how these mergers can be simulated in a familiar framework and to also show the range of welfare outcomes that are likely given this modeling environment.

Our model is drawn directly from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.  We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.  In each of these simulations, we calculate the impacts on consumer welfare.  Furthermore, we examine how the presence of additional integrated rivals (separate from the merging firms) changes our results.

Our simulations show certain patterns across mergers.  For horizontal upstream and downstream mergers, we find that consumers are net harmed in most instances.  The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be the largest.  Furthermore, we find that integrated mergers are typically net harmful to consumers as well.  The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.  Thus, although all of these mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.  Furthermore, we find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.  The existence of rivals with their own sources of bargaining leverage stemming from vertical integration appears to reduce the variance of merger outcomes.

@@

The paper proceeds as follows.  In Section \ref{sec:theory}, we introduce the theoretical framework and explain how we model mergers.  Section \ref{sec:sims} explains how we construct our numerical simulations and provides our results, and our application to Republic/Santek appears in Section \ref{sec:application}.  Additional simulation results appear in the Appendix.  Section \ref{sec:concl} concludes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Downstream and Upstream Competition}
Assume there is a set of consumers who each choose to buy a product sold by a single retailer.  Retailers are indexed by $r$, and the wholesalers that supply these retailers are indexed by $w$.  Prior to any mergers taking place, each wholesaler offers only one product, although each retailer can purchase from multiple wholesalers.  We denote the set of all retailers by $\mathbb{R}=\{1, \dots, \left\vert{\mathbb{R}}\right\vert\}$, and the set of all wholesalers by $\mathbb{W}=\{1, \dots, \left\vert{\mathbb{W}}\right\vert\}$.  We divide the set $\mathbb{W}$ into overlapping subsets, each labeled $\mathbb{W}^r$, to indicate which wholesalers' products are carried by which retailers.  Similarly, we divide the set of retailers $\mathbb{R}$ into overlapping subsets, each labeled $\mathbb{R}^w$, to indicate the retailers that carry the product sold by each wholesaler.

@@

Together the series of downstream and upstream first order conditions determine market equilibrium.  This model can be solved and calibrated as described in \citeN{ST2021}.  In our baseline configuration, we assume that wholesale and retail marginal costs are constant.  We allow for linear increasing marginal cost as a robustness check that appears in the Appendix.

@@

When the merged firm is bargaining with the unaffiliated wholesaler $v$ over what input price to pay, the bargaining first order condition becomes
\begin{equation}
\begin{split}
&[p^W_{rv} - c^W_{rv}]s_{rv} - \sum_{t \in \mathbb{R}^v \setminus \{r\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^r \setminus \{v\}) = \\
&\frac{1-\lambda}{\lambda} \left([p_{rv} - p^W_{rv} - c^R_{rv}]s_{rv} - \sum_{x \in \mathbb{W}^r \setminus \{w,v\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{v\})\right.\\
&\left.- \underbrace{[p_{rw} - c^W_{rw} - c^R_{rw}] \Delta s_{rw}(\mathbb{W}^r \setminus \{v\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{wholesale recapture leverage effect}}\right).
\end{split}
\label{eq: vmerger ret upstream foc}
\end{equation}
In this case, the merged firm has two channels for potential additional profits should it cease to trade with wholesaler $v$.  First, if retail sales are diverted to product $w$ sold by retailer $r$, those sales will earn a higher margin due to lower marginal costs stemming from what we call the ``EDM recapture effect.''  Second, the loss of product $v$ carried by retailer $r$ could increase sales by wholesaler $w$ through other retailers, which we call the ``wholesale recapture leverage effect.''  Both of these effects would tend to lower the resulting input price.

@@

As for wholesale prices, again consider a merger between an integrated retailer/wholesaler $rw$ and a standalone retailer $s$.  When wholesaler $w$ bargains with an unaffiliated retailer $u$ we have the first order condition given by,
\begin{equation}
\begin{split}
&[p^W_{uw} - c^W_{uw}]s_{uw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s,u\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^u \setminus \{w\})\\
& - \overbrace{\sum_{t \in \{r,s\}}\left(\overbrace{[p_{tw}-c^W_{tw}-c^R_{tw}] \Delta s_{tw}(\mathbb{W}^u \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^t \setminus \{w\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathbb{W}^u \setminus \{w\})\right)}^{\text{RRC effect}}=\\
&\frac{1-\lambda}{\lambda} \left([p_{uw} - p^W_{uw} - c^R_{uw}]s_{uw} - \sum_{x \in \mathbb{W}^u \setminus \{w\}} [p_{ux} - p^W_{ux} - c^R_{ux}] \Delta s_{ux}(\mathbb{W}^u \setminus \{w\})\right).
\end{split}
\label{eq: intdmerger wh upstream foc}
\end{equation}
The RRC effect is augmented with the profits emanating from the sales of retailer $s$, in addition to the sales of retailer $r$, both of which may potentially recapture sales should retailer $u$ lose access to product $w$.  The merged firm has higher bargaining leverage as a result.  If instead the integrated retailer/wholesaler $rw$ were to merge with a standalone wholesaler $v$, then the profits earned by firm $rw$ should the negotiation with retailer $u$ fail are augmented with the earnings of wholesaler $v$ rather than of retailer $s$.  This adds a term similar to the wholesale recapture leverage effect seen in equation \eqref{eq: vmerger ret upstream foc} to the left-hand side of the bargaining first order condition.\footnote{For simplicity, we assume that when a retailer fails to reach an agreement with wholesaler $w$, that retailer's contract with wholesaler $v$ remains in place.  This assumption can be loosened.  }    In the case of a merger between two integrated retailer/wholesaler firms, all of these effects would appear.

Returning to a merger between an integrated retailer/wholesaler $rw$ and a standalone retailer $s$, when the merged firm bargains with an unaffiliated wholesaler $v$ to supply retailer $r$, the first order condition becomes
\begin{equation}
\begin{split}
&[p^W_{rv} - c^W_{rv}]s_{rv} - \sum_{t \in \mathbb{R}^v \setminus \{r\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^r \setminus \{v\}) = \\
&\frac{1-\lambda}{\lambda} \left([p_{rv} - p^W_{rv} - c^R_{rv}]s_{rv} - \sum_{x \in \mathbb{W}^r \setminus \{w,v\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{v\})\right.\\
&- \underbrace{\sum_{t \in \{r,s\}}[p_{tw} - c^W_{tw} - c^R_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{EDM recapture effect}}
- \underbrace{\sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^r \setminus \{v\})}_{\text{retail recapture leverage effect}}\\
&-\left.\underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{wholesale recapture leverage effect}}\right).
\end{split}
\label{eq: intdmerger ret upstream foc}
\end{equation}
Compared to equation \eqref{eq: vmerger ret upstream foc}, now the EDM recapture effect applies to sales of product $w$ at both affiliated retailers $r$ and $s$.  Furthermore, there is also the possibility that sales will be diverted to retailer $s$ should retailer $r$ lose access to product $v$, which creates an additional ``retail recapture leverage effect'' alongside the wholesale recapture leverage effect.  All of these additional terms tend to increase the bargaining leverage of the merged firm.  If we instead examined a merger between retailer/wholesaler $rw$ and an unaffiliated wholesaler, that would augment the EDM recapture effect with sales of retailer $r$ for two merged wholesalers, rather than sales through two merged retailers.  Similarly, the retail recapture leverage effect would be replaced with an additional wholesale recapture leverage effect for the newly merged wholesaler.  For a merger between two integrated retailer/wholesaler firms, all of these effects would enter.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Simulations \label{sec:sims}}
In this section, we use a similar simulation setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Generating Process}
Broadly speaking, we consider four categories of mergers: downstream, upstream, vertical, and integrated.  We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.  Similarly, upstream mergers are those between two unintegrated wholesalers or between an unintegrated wholesaler and a vertically integrated wholesaler/retailer.  We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.  Finally, an integrated merger is between two firms that are both already vertically integrated pre-merger.

For downstream, upstream, and vertical mergers, we simulate markets by randomly sampling shares from a Dirichlet distribution for 2 to 5 retailers or wholesalers, respectively.\footnote{We parameterize the Dirichlet distribution so it is equivalent to a uniform distribution.  }  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers to 7. We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers). Vertically integrated firms are not siloed: integrated wholesalers supply inputs to retailers other than their integrated partner and vertically integrated retailers purchase inputs from wholesalers other than their integrated partner.

@@

The bargaining parameter ranges from from 0.1 (where wholesalers have the advantage) to 0.9 (where retailers have the advantage). We report our results in terms of relative bargaining power, $(1-\lambda)/\lambda$.  The bargaining parameter is identical for all of the retailers in each simulation.  We calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option.  All other goods are differenced relative to this option, resulting in the mean zero outside good normalization.  We set the market size to 1.

@@

Table \ref{tab:simsum} provides summary statistics for our simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.}  The 2010 DOJ/FTC Horizontal Merger Guidelines categorize ``Highly Concentrated Markets'' as those with Herfindahl Hirschman Index (HHI) values over 2,500 points.  The Guidelines state that mergers with HHI changes greater than 200 points that result in Highly Concentrated markets are ``presumed likely to enhance market power.''\footnote{See \S 5.3 of the Guidelines.  }  The vast majority of our mergers meet these conditions.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.  }  The median average wholesale pre-merger price is \$5.80, and the median average retail pre-merger price is \$15.  Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.

Figure \ref{fig:surplussum} summarizes our results for welfare.  Each of the four panels shows how mergers impact surplus for a particular set of agents (consumers, retailers, wholesalers, or the entire market combined, respectively).  Surplus is reported as a percentage change of total pre-merger expenditures in the downstream market.  Each panel contains four box and whisker plots, with each plot corresponding to a different type of merger. The whiskers show the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median. Negative outcome values imply harm, and positive values imply benefits.

@@

Throughout these simulations, we maintain the assumption that marginal costs are constant.  However, given that many of the mergers we study have vertical aspects where input cost considerations may drive the realization of EDM and RRC, it is reasonable to wonder how our results would change in an alternative marginal cost environment.  In the Appendix, we provide results that allow marginal costs to increase linearly.  We find that this alternative assumption tends to result in mergers that are more harmful for consumers and total welfare.

@@

We examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.  Figure \ref{fig:CVvertincumbBW_consumer} depicts box and whisker plots summarizing consumer welfare as the number of incumbent integrated firms increases. For the results on total welfare, see the Appendix.  The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated. By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties. For integrated mergers, the plot when the number of incumbent integrated firms equals 0 depicts a merger between two integrated firms, again with no integrated rivals.

@@

\begin{figure}
\centering
\includegraphics{output/TrashSimsCompare.png}
\caption{Comparison between Collection-Only and Full Vertical Simulations for Republic/Santek}
\label{fig:TrashSimsCompare}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Appendix: Increasing Marginal Costs}
